---
title: "constructing the components of a mixed model"
bibliography: "../glmm.bib"
csl: ../nature.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymbol}
- \newcommand{\X}{\boldsymbol X}
- \newcommand{\J}{\boldsymbol J}
- \newcommand{\Z}{\boldsymbol Z}
- \newcommand{\bbeta}{\boldsymbol \beta}
- \newcommand{\bb}{\boldsymbol b}
- \newcommand{\bu}{\boldsymbol u}
- \newcommand{\bLambda}{\boldsymbol \Lambda}
- \newcommand{\bEta}{\boldsymbol \eta}
- \newcommand{\btheta}{\boldsymbol \theta}
- \newcommand{\bzero}{\boldsymbol 0}
---


\newcommand{\bm}[1]{#1}

## back to the general definition

$$
\begin{split}
\underbrace{Y_i}_{\text{response}} & \sim \overbrace{\text{Distr}}^{\substack{\text{conditional} \\ \text{distribution}}}(\underbrace{g^{-1}(\eta_i)}_{\substack{\text{inverse} \\ \text{link} \\ \text{function}}},\underbrace{\phi}_{\substack{\text{scale} \\ \text{parameter}}}) \\
\underbrace{\boldsymbol \eta}_{\substack{\text{linear} \\ \text{predictor}}} & 
 = 
\underbrace{\X \boldsymbol \beta}_{\substack{\text{fixed} \\ \text{effects}}} + 
\underbrace{\Z \bb}_{\substack{\text{random} \\ \text{effects}}}
\\
\underbrace{\bb}_{\substack{\text{conditional} \\ \text{modes}}}  & 
\sim \text{MVN}(\boldsymbol 0, \underbrace{\Sigma(\btheta)}_{\substack{\text{variance-} \\ \text{covariance} \\ \text{matrix}}})
\end{split}
$$

## fixed-effect model matrix

* $\X$ ($n \times p_\textrm{fix}$)
* **terms**, **input variables**, **predictor variables**
* *Wilkinson-Rogers* notation [@WilkinsonRogers1973]
* defined by `model.matrix()`, `terms()` [@chambersStatistical1992a]
* details hard to find!
* default intercept term (`1`); `-1` or `+0` to suppress
* numeric input variables → single predictor variable
* variable transformations on the fly 
* *contrasts* automatically constructed for factors (unless `+0`)
* interaction `:` = pairwise products of predictor variables
* data-dependent bases: `poly()`, `splines::ns()`, etc.

## example

```{r eval=FALSE}
terms(~ mpg * disp, data = mtcars)
model.matrix(hp~ mpg * disp, data = mtcars)
```

## random-effect formulas

* formula: typically `( f | g )` (`f` = varying term, `g` = cluster/grouping variable)
* `g` can be expanded into separate terms (`(f | g1/g2)` → `(f | g1 + g1:g2)`); `*`, `+` *may* be expanded (`reformulas` package)
* `f` = any set of terms that vary *within* levels of `g`
* dialect differences
    * random component may be specified separately from fixed (e.g. `nlme::lme`, `MCMCglmm`)
	* ... or as part of the same formula (e.g. `lme4`, `glmmTMB`)
    * covariance structures may be specified (`brms`, `MCMCglmm`, `glmmTMB`) as (e.g.) `cs( f | g )`
	* AS-REML-like syntax (e.g. `MCMCglmm`): e.g. `~us(f):g`
	* extensions for multi-membership models, shared variables between terms, ...
* each RE term (after separation) is independent	
	
## from RE formula to model objects: RE model matrix $\Z$

* details in [lmer vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf) [@bates_fitting_2015]
* $\Z$
   * *Khatri-Rao product* of $\J$ (indicator matrix for cluster variable `g`) with model matrix for `f`
   * e.g.:
```{r k-r}
library(Matrix)
f <- factor(rep(1:3, each = 2))
Ji <- t(as(f, "sparseMatrix"))
Xi <- cbind(1, rep.int(c(-1, 1), 3L))
Zi <- t(Matrix::KhatriRao(t(Ji), t(Xi)))
```
* make $\J$ a **sparse** matrix, $\Z$ inherits structure
* concatenate $\Z$ from multiple terms columnwise

## from RE formula to model objects: RE covariance matrix $\Sigma$

* $\Sigma_i$ (for a single term) is block-diagonal (one block per cluster); terms are blocks too
* we may be interested in the lower-triangular *Cholesky factor* ($\Sigma = \Lambda \Lambda^\top$)
   * or the *precision matrix* ...
* `lme4`

## references {.refs}
