---
title: "correlation structures"
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymbol}
- \newcommand{\X}{\boldsymbol X}
- \newcommand{\J}{\boldsymbol J}
- \newcommand{\Z}{\boldsymbol Z}
- \newcommand{\bbeta}{\boldsymbol \beta}
- \newcommand{\bb}{\boldsymbol b}
- \newcommand{\bu}{\boldsymbol u}
- \newcommand{\bLambda}{\boldsymbol \Lambda}
- \newcommand{\bEta}{\boldsymbol \eta}
- \newcommand{\btheta}{\boldsymbol \theta}
- \newcommand{\bzero}{\boldsymbol 0}
---

\newcommand{\bm}[1]{#1}

## Review of RE covariance matrices ($\Sigma$)

* $\Sigma$ is doubly block-diagonal
   * one block for every random-effect *term* (maybe only 1)
   * one sub-block for every *cluster* within a term, maybe only 1 (R-side effects)
   
## Correlation structures in `lme`

- separate for R-side (`corStruct`: `corAR1`, `corCompSymm`, etc.), G-side `pdStruct`

```{r}
library(nlme)
data(Nile)
nile <- data.frame(date = 1871:1970, flow = c(Nile))
fit1 <- gls(flow ~ date, correlation = corAR1(), data = nile)
```

## R-side modeling

* structure of residual variance
* Ironically, most R packages don't do R-side structures specially
* Can construct an observation-level factor with the appropriate structure
* May need to set the actual residual variance term to zero (`glmmTMB`, `blme`)

## Correlation structures in `lme4`

* not much! strong assumption of unstructured covariance matrix
* `||` operator separates components into separate terms
* e.g. `(1 + x + y || f)` $\to$ `(1|f) + (0 + x | f) + (0 + y | f)`
* purely *semantic*; factor variables aren't handled well
* `afex::mixed()` adds a layer to produce a diagonal covariance matrix even for factors (by splitting/creating dummies)

## Correlation structures in `glmmTMB`: native

```{r}
library(glmmTMB)
nile$dummy <- factor(1)
fit2 <- glmmTMB(flow ~ date + ar1(factor(date) + 0 | dummy),
                data = nile,
                dispformula = ~ 0,
                REML = TRUE)
```

## Correlation structure via smooths

* `mgcv` smooths can (as of recently) be used in `glmmTMB`
  * may be rough around the edges
  * fewer downstream methods (inference, effective df) available
* also `gamm4` (but less convenient and flexible)
* constructs $\Z$, $\Sigma$ appropriately
* $\Sigma$ is always known up to a constant
  * e.g. scale, shape parameters for autoregressive/spatial terms must be specified *a priori*
  * less important if controlling for correlation rather than making inferences
  
## `mgcv` smooths

* long list: basic are different forms of 1D splines
* Gaussian process smooths
* **Markov random field** smooths (low rank approximation)

## Reduced-rank (factor-analytic) smooths

* factor-analytic models; 
* `rr(...)` in `glmmTMB`, `gllvm` package (more specialized)


## sdmTMB

* relatively new package
* uses meshes, stochastic PDE framework from INLA for sparsity

## spaMM

* appears capable ...

## pedigrees etc.

* I know a lot less about these!
* make $\Sigma$ proportional to the pedigree/phylogenetic matrix
* how to hack with `glmmTMB` (`map` function)

## refs {.refs}
