---
title: "modular fitting"
---

```{r knitr,include=FALSE}
require("knitr")
knitr::opts_knit$set(root.dir = here::here())
knit_hooks$set(crop=hook_pdfcrop)
```

```{r pkgs, message = FALSE}
library(lme4)
library(glmmTMB)
library(ggplot2); theme_set(theme_bw())
library(parameters)
library(mgcv)
```

## Goals

* allow more flexible constructions, work around package limitations
* e.g. work-in-progress [phyloglmm](https://zenodo.org/records/2639887), [multimembership models](https://rpubs.com/bbolker/groupmembers), examples from [lmer vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf) appendix

## General steps

1. construct `X`, `Z`, `Sigma` in appropriate form
2. construct objective function (negative log-likelihood/deviance)
3. optimize objective function
4. package results

**Modifications:** (a) alter components between steps 1 and 2; (b) write a wrapper function for the objective function and optimize it instead

* `glmer` uses a two-stage optimization (first optimizing over $\theta$ only, then $\{\theta, \beta\}$)

## `lme4` bare-bones modular example

```{r mod1}
lmod <- lFormula(Reaction ~ Days + (Days|Subject), sleepstudy)
names(lmod)
devfun <- do.call(mkLmerDevfun, lmod)
opt <- optimizeLmer(devfun)
fit <- mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
```

## example: AR1 correlation structure

From Wicklin, Rick (2018) [“Fast Simulation of Multivariate Normal Data with an AR(1) Correlation Structure.”](https://blogs.sas.com/content/iml/2018/10/03/ar1-cholesky-root-simulation.html), code to generate the Cholesky factor of an AR1 covariance directly.

```{r ar1cor}
ar1cor <- function(rho, p) {
    R <- matrix(0, p, p)
    R[1,] = rho^(0:(p-1))        ## formula for 1st row
    c <- sqrt(1 - rho^2)         ## scaling factor: c^2 + rho^2 = 1
    R2 <- c * R[1,]              ## formula for 2nd row
    for (j in  2:p) {            ## shift elements in 2nd row for remaining rows
        R[j, j:p] = R2[1:(p-j+1)] 
    }
    return(R)
}
ar1cor(0.8, 5)
devfun2 <- function(sigma, rho) {
    cholfac <- sigma*t(ar1cor(rho, p))
    theta <- cholfac[lower.tri(cholfac, diag = TRUE)]
    devfun(theta)
}
d <- (read.csv("data/Cod_daily_depth_data.csv")
    |> na.omit()
    |> transform(ctemp = Temperature_1m-mean(Temperature_1m),
                 fdate = factor(as.Date(date)))
)
```

```{r fit-ar1, cache = TRUE}
m1 <- glmmTMB(log(depth_mean_day) ~ 1 + ctemp +  (1 + ctemp | fish) +
            ar1(0 + fdate | fish),
        data = d)
m2 <- update(m1, dispformula = ~0)
AIC(m1, m2)
ggplot(d, aes(ctemp, -log(depth_mean_day))) + 
  geom_point(alpha = 0.3) +
  facet_wrap(~fish)
m1 <- lmer(log(depth_mean_day) ~ 1 + ctemp +  (1 + ctemp | fish), data = d) # exercise
## library(mgcv)
m2 <- gam(log(depth_mean_day) ~ s(ctemp, factor(fish), bs = "fs"), data = d)
   
## performance::check_model(m1)
## performance::check_model(m2)
## nloptwrap
```

## `glmmTMB` bare-bones modular example

```{r glmmTMB-1}
library(glmmTMB)
## construct components
m1 <- glmmTMB(count ~ mined + (1|site),
              family=poisson, data=Salamanders, doFit = FALSE)
## make TMB object (don't optimize)
m2 <- fitTMB(m1, doOptim = FALSE)
## optimize
m3 <- with(m2, nlminb(par, objective = fn, gr = gr))
## construct fitted model object
m4 <- finalizeTMB(m1, m2, m3)
```

## `glmmTMB`, rebuilding TMB object

If we modify the components of `m1$env$data` at this point ...
rebuild TMB structure (*may* be necessary)

```{r glmmTMB-2}
m2 <- with(m2$env,
           TMB::MakeADFun(data,
                          parameters,
                          map = map,
                          random = random,
                          silent = silent,
                          DLL = "glmmTMB"))
```

## use of the `map` argument in `glmmTMB`

* set parameters equal to starting values *or* fix values equal to each other
* e.g. for an AR1 model, `map = list(theta = c(1, NA)), start = list(theta = 0, 1)` would estimate the first parameter (log-SD) and fix the correlation to its starting value
* could pass a known correlation matrix (e.g. pedigree) by transforming into the `glmmTMB` parameterization and using `map`


